Translation of B Implementations to the LLVM: Control and Modularity

### 2013-09-23 translation.tex
Yet another work in progress version of the Python loader of BXML files. Able to load very basic operations, substitutions and expressions.
Profound changes in the text laying out the principles for composing translation of different modules related by import clauses. The specification of the translation has not been updated accordingly and is deprecated.

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=false]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\pagestyle{headings}
\title{Translation of B Implementations to the LLVM: Control and Modularity}
\author{David Déharbe \and Valério Medeiros Jr} 
\institute{Program of Graduate Studies in Systems and Computing (PPgSC) \\
  Federal University of Rio Grande do Norte \\
  Natal (Brazil)}
\date{May 2013}

@@

        \B{see} & : & seq \B{Mach} \\
        \B{impo} & : & seq \B{Impo} \\

@@

        \multicolumn{3}{|c|}{Imported machine: \B{Impo}} \\

@@

        \B{mach} & : & \B{Mach} \\

@@

        \multicolumn{3}{|c|}{Machine: \B{Mach}} \\

@@

        \B{impl} & : & \B{Impl} \\

@@

Local entities are constants, variables or functions. An example of constant
definition is \llvm{@secret = constant i32 42}, it is composed of a name, type
and value. The elements of a variable definition are the name, type and code
generation attributes, e.g \llvm{@count = common global i32 zeroinitializer}.
Attributes \llvm{common}, \llvm{global} and \llvm{zeroinitializer} provide
information for target code generation: linkage type (the variable is merged
with other of the same name), the scope (global) and initialization of the
variable (zeroed is mandatory for common linkage). LLVM defines other attributes
for these aspects, but they are not needed in this work and their presentation
is omitted.  Function definitions are composed of the signature and body. The
signature contains the return type, name, parameters, and attributes for target
code generation. Attribute \llvm{nounwind} states that the function never
generate exceptions. The body is a sequence of blocks of instructions in
single-static assignment (SSA) form, i.e. are such that each variable may be
assigned exactly once.

The grammar rule \nt{inst} describes the different kind of
instructions. Instruction \llvm{alloca} allocates a memory block the size of the
given type from the stack segment. This and the following classes of
instructions assign the resulting value to a fresh variable. Arithmetic
operations are binary, and include signed quotient and reminder. Arithmetic
comparisons compare two values according to the first operand and return a 1-bit
integer value.  Instruction \llvm{call} invokes the given function with the
given arguments and assigns the result to a fresh variable, which is of the
given type.  In general, \llvm{getelementptr} gets the address of an element in
an aggregate object through indexing. In the target LLVM subset, aggregate
objects are structures, the first index has type \llvm{i32} and value \llvm{0}
to select the first structure at the given location \nt{exp}, and the second
index selects a field in that structure.  Instruction \llvm{load} assigns to a
fresh variable \nt{name} the contents of a memory address of type \nt{type}
specified by \nt{exp} (e.g. in fig.~\ref{fig:ex-llvm}).  Instruction
\llvm{store} writes a value to memory address (again, see
fig.~\ref{fig:ex-llvm}).  Instruction \llvm{br} has two forms.  In the first
form, it directs conditionally the control flow to one of two blocks.  If the
given expression evaluates to \llvm{1}, control goes to the first block,
otherwise to the second. In the second form, the control jumps unconditionally
to the given block.
% Instruction \llvm{switch} directs the
% control flow to one of several blocks. The given expression is evaluated and
% control goes to the first branch with the matching value, or to the default
% branch (appearing first in the instruction).  
Finally, instruction \llvm{ret}
ends the current function call, optionally returning a value.

@@

\section{Translation: a Bird's-Eye View}
\label{sec:overview}

The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
module, say \llvm{M}. A B implementation \B{M} has several sections, each
translated as follows:
\begin{enumerate}
\item Visibility clauses provide access to definitions found in external
  B components and are translated to external declarations in LLVM.
\item Constants and value definitions are translated to their value in the
  LLVM platform.
\item To encode the implementation state, the translation produces a structure
  type, named \llvm{\%M\$state\$}, and an instance of that type, named
  \llvm{@M\$self\$}. All the variables and instances of imported machines are
  encoded as elements of that structure. These elements are referenced by their
  position in the structure within the LLVM code. Also, for each operation of
  each imported machine, an external function declaration is produced.
\item Each operation \B{op} of the implementation is translated into the
  definition of a function named \llvm{@M\$op}; also the translation produces a
  special function \llvm{@M\$init\$} responsible for the initialization.

  The parameter list of the functions thus produced contains one item for each
  input and output of the corresponding operation, and one item of type
  \llvm{\%M\$state\$*}, to pass the address of the structure encoding the
  state of the implementation and its modification by the callee. The return
  type is always \llvm{void}.
\end{enumerate}
The translation is designed to support modular B development. In a B
implementation, one may reuse existing components, by instantiating (the B term
is \emph{importing} so-called library machines. Such library machines have their
own development and implementation. Borrowing terminology from the
object-oriented world, one may view library machines as classes, and the B
implementation as an object that contains instances of those classes. In
practice, different from most OO languages, such instances are not references,
but the actual objects. We consider this necessary as dynamic memory allocation
is precluded.

To translate the instantiations of library machines in an implementation, the
translator needs to have access to the following information on those machines
and their respective implementation:
\begin{itemize}
\item for each library machine \B{M}, the corresponding implementation, denoted
  \B{M.impl}.
\item for such implementation, the type of the state and the signature of the
  operations.
\end{itemize}
Note that, when translating the implementation of a library machine, no instance
shall be generated. It is in the implementation of the main design, that we call
the \emph{top-level} component, that such instances are created.

Figure~\ref{fig:skel} summarizes the translation principles by providing the
structure of the LLVM implementation thus produced.

\begin{figure}
  \begin{center}
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{for each imported component} \B{Q} \textit{generate} \\
      \> \llvm{\%Q\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \> \llvm{declare void @Q\$init\$(\%Q\$state\$*)} \\
      \> \textit{for each operation} \B{op} \textit{generate} \\
      \> \> \llvm{declare void @Q\$op(\%Q\$state\$*, \ListOf{\nt{type}})} \\
      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
      \llvm{define void @M\$init\$(\%M\$state\$* \%self\$) \{} \\
      \> \llvm{\ListOf{\nt{block}}} \\
      \> \llvm{exit: ret void} \\
      \llvm{\}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
      \> \> \llvm{\ListOf{\nt{block}}} \\
      \> \> \llvm{exit: ret void} \\
      \> \llvm{\}} \\
      \textit{if this is the top-level component, generate} \\
      \> \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer}
    \end{tabbing}
  \end{center}
  \caption{Structure of the LLVM module produced from a B implementation \B{M}.}
  \label{fig:skel}
\end{figure}

@@

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=false]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
\pagestyle{headings}
\title{Translation of B Implementations to the LLVM: Control and Modularity}
\author{David Déharbe \and Valério Medeiros Jr} 
\institute{Program of Graduate Studies in Systems and Computing (PPgSC) \\
  Federal University of Rio Grande do Norte \\
  Natal (Brazil)}
\date{May 2013}

@@

        \B{impo} & : & seq \B{Module} \\

@@

        \multicolumn{3}{|c|}{Modules: \B{Module}} \\

@@

        \B{mach} & : & \B{MachRef} \\

@@

        \multicolumn{3}{|c|}{Machine: \B{MachRef}} \\
        \B{spec} & : & \B{MachInterface} \\
        \B{impl} & : & opt \B{Impl} \\
        \multicolumn{3}{|c|}{Machine interface: \B{MachInterface}} \\

@@

@@

Local entities are constants, variables or functions. An example of constant
definition is \llvm{@secret = constant i32 42}, it is composed of a name, type
and value. The elements of a variable definition are the name, type and code
generation attributes, e.g \llvm{@count = common global i32 zeroinitializer}.
Attributes \llvm{common}, \llvm{global} and \llvm{zeroinitializer} provide
information for target code generation: linkage type (the variable is merged
with other of the same name), the scope (global) and initialization of the
variable (zeroed is mandatory for common linkage). LLVM defines other attributes
for these aspects, but they are not needed in this work and their presentation
is omitted. For each such definition, a memory block is statically allocated to
stores the variable value. Function definitions are composed of the signature
and body. The signature contains the return type, name, parameters, and
attributes for target code generation. Attribute \llvm{nounwind} states that the
function never generate exceptions. The body is a sequence of blocks of
instructions in single-static assignment (SSA) form, i.e. are such that each
variable may be assigned exactly once.

The grammar rule \nt{inst} describes the different kind of instructions. All
instructions producing a value assign it to a fresh variable.  Instruction
\llvm{alloca} allocates a memory block the size of the given type on the stack
segment. This memory is automatically freed when the current frame is popped
from the stack. The LLVM does not include instructions for allocating from the
heap. Arithmetic operations are binary, and include signed quotient and
reminder. Arithmetic comparisons compare two values according to the first
operand and return a 1-bit integer value.  Instruction \llvm{call} invokes the
given function with the given arguments and assigns the result to a fresh
variable, which is of the given type.  In general, \llvm{getelementptr} gets the
address of an element in an aggregate object through indexing. In the target
LLVM subset, aggregate objects are structures, the first index has type
\llvm{i32} and value \llvm{0} to select the first structure at the given
location \nt{exp}, and the second index selects a field in that structure.
Instruction \llvm{load} assigns to a fresh variable \nt{name} the contents of a
memory address of type \nt{type} specified by \nt{exp} (e.g. in
fig.~\ref{fig:ex-llvm}).  Instruction \llvm{store} writes a value to memory
address (again, see fig.~\ref{fig:ex-llvm}).  Instruction \llvm{br} has two
forms.  In the first form, it directs conditionally the control flow to one of
two blocks.  If the given expression evaluates to \llvm{1}, control goes to the
first block, otherwise to the second. In the second form, the control jumps
unconditionally to the given block.
% Instruction \llvm{switch} directs the
% control flow to one of several blocks. The given expression is evaluated and
% control goes to the first branch with the matching value, or to the default
% branch (appearing first in the instruction).  
Finally, instruction \llvm{ret}
ends the current function call, optionally returning a value.

@@

\section{Translation: a global picture}
\label{sec:overview}

In addition to producing LLVM code that is faithful to the design developed in
B, our code generator is designed to satisfy the following two requirements:
\begin{enumerate}
\item Memory is allocated statically, save for the function call frame stack: no
  dynamic memory allocation library is required. Many safety critical systems
  preclude use of this programming facility.
\item The code generator features separate compilation: an internal change in a
  component should only require generating new LLVM code for that component, and
  not for the other components that depend on it. This condition is important
  for large, distributed projects.
\end{enumerate}

In Atelier-B, software development is organized in libraries of modules. A
module has a specification, called a machine. A machine may be:
\begin{description}
\item[basic] In that case its implementation is carried out outside the realm of
  the B method. A possible application is a set of related I/O routines. Even
  though no LLVM code needs to be generated to implement the basic machines, the
  code generator needs still to produce LLVM declarations of the entities
  (e.g., constants and operations) that are supplied by those machines.
\item[developed] In that case, this module also contains a series of
  refinements, eventually reaching an implementable version, from which LLVM
  code may be generated.
\end{description}
Amongst the several constructs of the B notation supporting modular design, the
\emph{import} relation is related to separate compilation. At the implementation
level, one module may import several instances of a machine to construct its
internal data structures. Such imported machines may be either basic, or
developed. In the latter case, the implementation of the imported machines may
in turn import other machine instances and there is no pre-established limit to
such chain of imports. Hence, the import relation between module instances forms
a tree, where the root is an implementation and the descendants of a node are
the instances imported from the module in that node.

\begin{figure}
\begin{center}
\includegraphics[width=.8\textwidth]{b-project.pdf}
\caption{Structure of the imports relation in a B project}
\label{fig:structure}
\end{center}
\end{figure}
Figure~\ref{fig:structure}
illustrates this with the structure of a B project where an implementation
\B{I0} of a specification \B{M0} imports one unnamed instance of a basic machine
\B{M2} and two instances named \B{A} and \B{B} of a developed machine \B{M1}. In
turn its implementation \B{I1} itself imports two instances \B{C} and \B{D} of a
developed machine \B{M3}, implemented as \B{I3} (which has no imports itself).
The generation of LLVM code for the project shall allocate memory for one
instance of \B{I0}, two instances of \B{I2} and four instances of \B{I3}.

At that point, let us remark that, when generating code for a given module, we
need to consider the following two cases:
\begin{enumerate}
\item The code generation aims at producing a component, that will further be
  instantiated by importations. We call this the \emph{COMP} mode for code
  generation. This case is further split into two sub-cases:
  \begin{enumerate}
  \item The module is basic, and only declarations need to be emitted.
  \item The module is developed, and declarations and definitions are produced.
  \end{enumerate}
\item The generated code corresponds to a full project. Given the root module,
  the code generator must identify all the transitively imported module instances. We call this the \emph{PROJ} mode for code generation.
\end{enumerate}
One key issue in the code generation is how memory shall be allocated to store
the state of the modules. To satisfy our first initial requirement (no dynamic
memory allocation), this allocation shall be static, that is to say the
generator must include the code that creates the module instances when the
project starts execution. Note that this cannot be done in \emph{COMP} mode as,
at that point, we cannot assume any knowledge on the number of instances the
module being compiled we there be at run time. Consequently, the responsibility
for the static allocation of instances is assigned to the \emph{PROJ} mode,
which shall also perform the proper bindings between the actual and formal
instances in the generated code.

To encode the data space of a module, the translation produces a structure type,
named \llvm{\%M\$state\$}, such that the elements of the structure represent the
variables and the instances of imported modules. To satisfy our second initial
requirement (separate compilation), the instances cannot be part of the
structure itself, instead references to those instances (encoded as pointers)
are used. Let \llvm{\%M\$ref\$} denote the type pointer to \llvm{\%M\$state\$}.
To encode behavior of a module, each operation \B{op} is translated into the
definition of a function named \llvm{@M\$op}The parameter list of the functions
thus produced contains one item for each input and output of the corresponding
operation. There is also one parameter of type \llvm{\%M\$ref\$}, to pass the
address of the structure encoding instance associated with that operation. Also
the translation produces a special function \llvm{@M\$init\$} responsible for
the initialization. The parameters of this function are the addresses of the
instances found in the import tree of the module (including the module itself).

Our approach consists in dividing the translation into three sections, and to
combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
a given module \B{M}, the sections are the following:
\begin{description}
\item[typedef:] Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
  the data space of the module. In case the data space of a module is empty,
  this section is empty and the type is not defined:
  \begin{tabbing}
    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
  \end{tabbing}
\item[interface:] Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data}, if
  this has been defined. 
  \begin{tabbing}
    \llvm{\%M\$ref\$ = type \%M\$state\$*}
  \end{tabbing}
  The interface also contains declarations of LLVM
  functions: one implementing the initialisation of the module, and one for each
  operation of the module. If the data space of the module is empty, no function
  definition corresponding to the initialisation is emitted
  \begin{tabbing}
    foo \= \kill
    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
\item[implementation:] Defines the functions declared in the interface. This
  section is produced only for the developed modules.
  \begin{tabbing}
    foo \= \kill
    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
    \> \llvm{\ListOf{\nt{block}}} \\
    \> \llvm{exit: ret void} \\
    \llvm{\}}
  \end{tabbing}
\end{description}
Assuming these three sections are available, the \emph{COMP\/} translation
assembles them as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each imported module} \B{Q} \textit{generate} \\
    \> \textit{generate a type declaration for the state space of} \B{Q} \\
    \> \llvm{\%Q\$state\$ = type opaque} \\
    \> \textit{include the interface section of }\B{Q} \\
    \textit{include the typedef section of} \B{M} \\
    \textit{include the implementation section of} \B{M}
  \end{tabbing}
\end{center}
For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each transitively imported component} \B{Q} \\
    \> \textit{include the typedef section of \B{Q}} \\
    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\

    \textit{for each imported component} \B{Q} \\
    \> \textit{include the interface section of} \B{Q} \\
    \textit{define a function} \llvm{\%\$init\$} \textit{with a call to the } \\
    \> \textit{initialization function of} \B{M} \textit{with the proper bindings} \\
    \> \llvm{define void @\$init\$(void) \{} \\
    \> \> \llvm{call void @M\$init\$(@M, \{ \ListOf{instances} \}) \{} \\
    \> \> \llvm{exit: ret void} \\
    \> \llvm{\}} \\
    \> 
  \end{tabbing}
\end{center}

@@

[[[ THIS SECTION IS DEPRECATED ]]]

@@

[[[ DEPRECATED ]]]
